Growing Up with Nell: A Narrative Interface for Literacy

### 2012-03-12 paper.tex
Final copy editing, tweaking line breaks, removing overful hboxen, etc.

% THIS IS SIGPROC-SP.TEX - VERSION 3.1
% WORKS WITH V3.2SP OF ACM_PROC_ARTICLE-SP.CLS
%
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission,
% you need to 'insert'  your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%\documentclass{acm_proc_article-sp}
\documentclass[preprint]{sig-alternate}
\usepackage{url}
\usepackage{ifthen}\newboolean{noauthors}\setboolean{noauthors}{true}

@@

She nods and listens to a story about a princess.  Later, when
the Rahel has learned a little more, she will tell the machine that the
princess is named ``Rahel'' like she is and that she likes to wear blue---but
for now the green book draws pictures of the unnamed Princess for
her and asks her to trace shapes on the screen.

@@

Children (like all humans) are hard-wired for stories~\cite{boyd:stories}.
In our earlier learning software, we've seen some of our favorite
pedagogical activities neglected because children had difficulty
finding their way into the material without an enthusiastic teacher's
guidance.  In response, Nell uses a
\textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative}
to pave a path for the child user through the pedagogical content.
In this interface, all system actions are shaped by a storybook metaphor, and
interactions with the system are conceptualized as interactions with one of the
system protagonists.

@@

Children learn in different ways.  Nell provides multiple alternative
story modules for its plot points/lessons to engage the child's
interests and cater to multiple intelligences~\cite{gardner:mi}.
Selecting between the alternatives can depend on either explicit
choices made by the child (``fractions in outer space,'' if the child
chose a space-themed story or previously indicated an interest in space)
or prior success with a given
lesson style (a large number of accomplishments in musical-rhythmic
tasks suggests taking a rhythmic approach to fractions).  Similarly, several
different achievements can coexist, rewarding different ways of
accomplishing the same pedagogical goal.

This record of past choices and accomplishments is stored in a
\textit{Journal} and can be reviewed by the child.  The contents of the
Journal can be rearranged to create a Portfolio~\cite{stefanakis:portfolios}
demonstrating the child's progress.
In order to encourage fearless play and experimentation, the Journal
also supports pervasive undo; the child can rewind and replay the
narrative starting at any past point in the journal.

Content can also be remixed for further personalization.
The child is encouraged to rename characters and change clothing,
colors, and other superficial details.  In the future
we expect to allow further narrative customization by
recombination of story elements.
Lebowitz~\cite{lebowitz:universe85} and Riedl~\cite{riedl:planning}
show how a planner can be used to recombine and adapt story fragments.
We have less ambition than the cited work: instead of
attempting to generate thousands of stories from tens of templates, we
hope to select and then modestly adapt from hundreds of story
modules created in a decentralized manner by teachers---and eventually
by the students themselves.

@@

Second, Nell seeks to grow its authoring community.  Decentralized
authorship ensures that Nell's instructional content
will continue to grow, with an increasing number of topics and
customizations.  There are no
ceilings barring a child from participating: Nell
contains a story editor and everything necessary to
author and publish story modules.

To further promote collaboration
Nell is implemented in standard Web technologies (HTML5,
WebGL, and JavaScript) with offline caching.
Resources are named by URL, even when disconnected from the
internet, which simplifies the distribution of updates to story modules and
the Nell system.  URL-based naming allows third parties
to manage their own namespaces when extending Nell.

Finally, Nell is designed to allow a capable child to learn
the construction of the system itself,
eliminating the ceiling between the child and
the authors of the Nell system.  Nell can teach
about itself, and its source code is open for exploration within Nell.
Meaningful changes can be performed without external tools.

@@

\section{Technology}
Our technology choices further our goals of modularity and
decentralization.  Nell's implementation is free and open source and uses
standard web technologies to broaden the contributor base.  In this
section we describe the technologies that enable Nell's four key ideas.

\subsection{Rule-based story modules}
\begin{figure}\small
\begin{verbatim}
{ world: { // id gives canonical URL for this module
    id: "/nell.laptop.org/chapters/alphabet-book",
    scene: "opening", // initial scene
    inherit_from: "/nell.laptop.org/core/drawing_activity",
  }, scenes: [
    { id: "opening", subject: "actors/princess",
      verb: "speak", object: "actors/user",
      text: "Hello! Would (o/2) like to hear a story?",
      group: "/nell.laptop.org/core/story-intro-group",
    },
    { id: "story-1",
      cond: rule`At(opening) && Event(Chat, Yes)`,
      subject: "actors/princess",
      verb: "speak", object: "actors/user",
      text: "Once upon a time..." },
    ...
    { id: "r-drawn",
      cond: rule`Event(Activity, DrawR)`,
      subject: "actors/princess",
      verb: "speak", object: "actors/user",
      text: "Very good, (o/3)!",
      action: js`startAnimation("r-runs")` }, ...
  ], ... }
\end{verbatim}
\caption{Excerpt of story module.  JSON syntax has been extended with
quasiquote~\cite{quasiquote}.}\label{fig:rules}
\end{figure}
To implement Nell's modular \textit{narrative interface}, we use a
production rule system with programmable conflict resolution.
Figure~\ref{fig:rules} shows
an excerpt of the story module underlying the interaction described in
Section~\ref{sec:lesson}.

@@

Conflicts are resolved by consulting a \textit{group} associated with
each scene.  Each group has a default priority and names a scene to be
invoked when a conflict exists; the actions associated with that scene
will (eventually) alter the priorities of the conflicting scenes to
resolve the conflict.  The resolution scene may include any number of
actions.  It may select randomly among the conflicting scenes or base
its selection on
child preferences, curriculum progress, or inferred learning style.
The resolution scene may even initiate a new multi-scene story.

For example, the \textit{story-intro-group}
used in Figure~\ref{fig:rules} defaults to a low priority so that
continuing a story in progress is preferred over starting a new story.
When a story module is completed, several new story openings will be in
conflict.  The resolution scene for \textit{story-intro-group} brings us
to the controls of a magic airship which flies between locations
corresponding to the various story continuations.  In that set of
scenes, the child may eventually decide to fly to Mathland to hear the
next episode in her agent Mathis' adventure, which would bump up the
appropriate episode's priority to allow us to make the transition
without further conflict.

Nell will also include an integrated story editor.  Our exemplar is
the Wide Ruled authoring tool, which was successfully used by
non-technical story creators~\cite{skorupski:2009}.

@@

\subsection{Extensible dialog}
Nell's use as an interactive diary and
Portfolio~\cite{stefanakis:portfolios} is enhanced by the user's
collaboration in the fiction that Nell is intelligent---what
Turkle~\cite{turkle:alone} calls the ``ELIZA effect.''
Convincing discourse is a hard problem, but
reasonable approximations do not have to be difficult; even
the rudimentary conversational abilities of ELIZA elicited hours of
conversation~\cite{weizenbaum:power}.
% her secretary is divulging her life story
% asks him to leave the room so he could be alone with it
% hours discussing her boyfriend.
We've chosen to allow the child to directly converse with the
Princess and other agents within Nell.

@@

Figure~\ref{fig:aiml}
shows an AIML fragment augmenting an agent with
commands suitable for a text adventure story module.
In such a story the child might write instructions such as
\texttt{go north}, \texttt{go east}, or \texttt{open door} in
addition to the usual dialog with the agent.
These new commands fire events which are referenced by scene
preconditions.  For example, \texttt{north} may
trigger a scene which causes the activity to draw a new location on
the page, causes the agent to write and speak a description of the new
location, and which causes a shift in the AIML \textit{topic} to
enable additional vocabulary suitable for the new location.

@@

We have developed a block-oriented view of JavaScript, which we call
TurtleScript~\cite{turtlescript}.
The TurtleScript view of JavaScript is the single
implementation language for Nell.  TurtleScript scales from Logo-like
pedagogical tasks (in subsets of the language) up to the
implementation of TurtleScript and the
Nell system itself.  As Figure~\ref{fig:turtlescript} shows,
TurtleScript's block view is isomorphic to
JavaScript's standard textual
representation to preserve the code-reading skills of existing
programmers.  We have begun implementation of a
tablet programming environment for TurtleScript which uses
drag-and-drop, direct interaction, and intelligent prompting to reduce
text entry requirements.

@@

The TinkRBook system~\cite{chang:tinkrbook} shares a similar focus on
personalization and literacy.  Nell attempts to scale up these
ideas with a modular extensible narrative system and a broader range
of pedagogic material.

The Sugar Learning Platform~\cite{sugar} is an earlier effort at
child education which we've been directly involved in.  Nell's focus is
on children even further from educational infrastructure than those
targeted by Sugar.
The narrative at Nell's core attempts to provide students additional
pedagogical guidance, and the direct interaction model avoids keyboard
problems which have plagued Sugar's initial deployments.  Nell also
attempts to provide a greater sense of ownership through pervasive
customization.

@@

Automatic narrative generation has seen attention in
gaming, simulation, and training contexts~\cite{riedl:2008}.
We believe
Nell's use of these ideas in a pedagogical context is unique.

\section{Conclusions}
We have described the design of a novel narrative direct-interface
system for teaching literacy, which is personalized for and grows with
its child owner.  Personalized direct interface engages the child, and
narrative guides them through the pedagogic material.  The low-cost
solar-powered deployment hardware allows Nell to reach further into the
least-developed areas of the world to help those without traditional
educational infrastructure.

>>>>

% THIS IS SIGPROC-SP.TEX - VERSION 3.1
% WORKS WITH V3.2SP OF ACM_PROC_ARTICLE-SP.CLS
%
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission,
% you need to 'insert'  your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%\documentclass{acm_proc_article-sp}
\documentclass[preprint]{sig-alternate}
\usepackage{url}
\usepackage{ifthen}\newboolean{noauthors}\setboolean{noauthors}{true}
\hyphenation{Web-GL Java-Script Turtle-Script infra-structure}

@@

She nods and listens to a story about a princess.  Later, when
the girl has learned a little more, she will tell the machine that the
princess is named ``Rahel'' like she is and that she likes to wear blue---but
for now the green book draws pictures of the unnamed Princess for
her and asks her to trace shapes on the screen.

@@

Children (like all humans) are hard-wired for stories~\cite{boyd:stories}.
In our earlier learning software, we've seen some of our favorite
pedagogical activities neglected because children had difficulty
finding their way into the material without an enthusiastic teacher's
guidance.  In response, Nell uses a
\textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative}
to pave a path for the child user through the learning material.
In the narrative interface, all system actions are shaped by a
storybook metaphor, and
interactions with the system are framed as interactions with one of the
system protagonists.

@@

Children learn in different ways.  Nell provides multiple
story modules for its plot points/lessons to engage the child's
interests and cater to multiple intelligences~\cite{gardner:mi}.
Selecting between the alternatives can depend on either explicit
choices made by the child (``fractions in outer space,'' if the child
chose a space-themed story or previously indicated an interest in space)
or prior success with a given
lesson style (a large number of accomplishments in musical-rhythmic
tasks suggests a rhythmic approach to fractions).  Similarly, several
different achievements can coexist, rewarding different ways of
accomplishing the same pedagogical goal.

The record of past choices and accomplishments is stored in a
\textit{Journal} and can be reviewed by the child.  The contents of the
Journal can be rearranged to create a Portfolio~\cite{stefanakis:portfolios}
demonstrating the child's progress.
In order to encourage fearless play and experimentation, the Journal
also supports pervasive undo; the child can rewind and replay the
narrative starting at any past point in the journal.

Content can be remixed for further personalization.
The child is encouraged to rename characters and change clothing,
colors, and other superficial details.  In the future
we expect to accomplish further narrative customization by
recombination of story elements.
Lebowitz~\cite{lebowitz:universe85} and Riedl~\cite{riedl:planning}
show how a planner can be used to recombine and adapt story fragments.
We have less ambition than the cited work: instead of
attempting to generate thousands of stories from tens of templates, we
hope to select and then modestly adapt from hundreds of story
modules created in a decentralized manner by teachers---and eventually
by the students themselves.

@@

Second, Nell seeks to grow its authoring community.  Decentralized
authorship ensures that Nell's instructional content
continues to increase its topics and
customizations.  There are no
ceilings between children and teachers: Nell
contains a story editor and everything necessary to
author and publish story modules.

To further promote collaboration,
Nell is free and open source and implemented in standard web
technologies (WebGL, HTML5, and JavaScript) with offline caching.
Resources are named by URL, even when disconnected from the
internet, which simplifies the distribution of updates to story modules and
the Nell system.  URL-based identifiers also allow third parties
to manage their own namespaces when extending Nell.

Finally, Nell is designed to allow a capable child to learn about
the construction of the system itself,
eliminating the ceiling between the child and
the authors of the Nell system.  Nell can teach
about itself, and its source code is open for exploration within Nell.
Meaningful changes can be performed without external tools.

@@

\section{Technology}
%Our technology choices further our goals of modularity and
%decentralization.  Nell's implementation is free and open source and uses
%standard web technologies to broaden the contributor base.  In this
%section we describe the technologies that enable Nell's four key ideas.

Our implementation choices further our goals of modularity and
decentralization.  In this
section we describe the technologies that enable Nell's four key ideas.

\subsection{Rule-based story modules}
\begin{figure}\small
\begin{verbatim}
{ world: { // id gives canonical URL for this module
    id: "/nell.laptop.org/chapters/alphabet-book",
    scene: "opening", // initial scene
    inherit_from: "/nell.laptop.org/core/drawing_activity",
  }, scenes: [
    { id: "opening", subject: "actors/princess",
      verb: "speak", object: "actors/user",
      text: "Hello! Would (o/2) like to hear a story?",
      group: "/nell.laptop.org/core/story-intro-group",
    },
    { id: "story-1",
      cond: rule`At(opening) && Event(Chat, Yes)`,
      subject: "actors/princess",
      verb: "speak", object: "actors/user",
      text: "Once upon a time..." },
    ...
    { id: "r-drawn",
      cond: rule`Event(Activity, DrawR)`,
      subject: "actors/princess",
      verb: "speak", object: "actors/user",
      text: "Very good, (o/3)!",
      action: js`startAnimation("r-runs")` }, ...
  ], ... }
\end{verbatim}
\caption{Excerpt of story module.  JSON syntax has been extended with
quasiquote~\cite{quasiquote}.}\label{fig:rules}
\end{figure}
To implement Nell's modular narrative interface, we use a
production rule system with programmable conflict resolution.
Figure~\ref{fig:rules} shows
an excerpt of the story module underlying the interaction described in
Section~\ref{sec:lesson}.

@@

Conflicts are resolved by consulting a \textit{group} associated with
each scene.  Each group has a default priority and names a scene to be
invoked when a conflict exists; the actions associated with that scene
will (eventually) alter the priorities of the conflicting scenes to
resolve the conflict.  The resolution scene may include any number of
actions.  It may select randomly among the conflicting scenes or base
its selection on the
child's preferences, curriculum progress, or inferred learning style.
The resolution scene may even initiate a new multi-scene story.

For example, the \textit{story-intro-group}
used in Figure~\ref{fig:rules} defaults to a low priority so that
continuing a story in progress is preferred above starting a new story.
When a story module is completed, several new story openings will be in
conflict.  The resolution scene for \textit{story-intro-group} may bring us
to the controls of an airship which flies between locations
corresponding to the various story continuations.  In that set of
scenes, the child may eventually decide to fly to Mathland to hear the
next episode in her agent Mathis' adventure, which would bump up the
appropriate scene's priority to allow us to make the transition
without further conflict.

Nell will include an integrated story editor.  Our exemplar is
the Wide Ruled authoring tool, which was successfully used by
non-technical story creators~\cite{skorupski:2009}.

@@

\subsection{Extensible dialog}
Nell's use as an interactive diary and
\textit{portfolio}~\cite{stefanakis:portfolios} is enhanced by the child's
collaboration in the fiction that Nell is intelligent---what
Turkle~\cite{turkle:alone} calls the ``ELIZA effect.''
Convincing discourse is a hard problem, but
reasonable approximations do not have to be difficult; even
the rudimentary conversational abilities of ELIZA elicited hours of
conversation~\cite{weizenbaum:power}.
% her secretary is divulging her life story
% asks him to leave the room so he could be alone with it
% hours discussing her boyfriend.
We've chosen to allow the child to directly converse with the
Princess and other agents within Nell.

@@

Figure~\ref{fig:aiml}
shows an AIML fragment augmenting an agent with
commands suitable for a text adventure story module.
In such a story the child might write instructions such as
\texttt{go north}, \texttt{go east}, or \texttt{open door} in
addition to the usual dialog with the agent.
These new commands fire events which are referenced by scene
preconditions.  For example, \texttt{north} may
trigger a scene which causes the activity to draw a new location on
the page, causes the agent to write and speak a description of the new
location, and finally causes a shift in the AIML \textit{topic} to
enable additional vocabulary suitable for the new location.

@@

We have developed a block-oriented view of JavaScript, which we call
TurtleScript~\cite{turtlescript}.
The TurtleScript dialect of JavaScript is the single
implementation language for Nell.  TurtleScript scales from Logo-like
pedagogical tasks (in subsets of the language) up to the
implementation of TurtleScript and the
Nell system itself.  As Figure~\ref{fig:turtlescript} shows,
TurtleScript's block view is isomorphic to
JavaScript's standard textual
representation to preserve the code-reading skills of existing
programmers.  We have begun implementation of a
tablet programming environment for TurtleScript which uses
drag-and-drop, direct interaction, and intelligent prompting to reduce
text entry requirements.

@@

The TinkRBook system~\cite{chang:tinkrbook} shares a similar focus on
personalization and literacy.  Nell attempts to scale the system up
with a modular extensible narrative system and a broader range
of pedagogic material.

The Sugar Learning Platform~\cite{sugar} is an earlier effort at
child education in which we've been directly involved.  Nell's focus is
on children further from educational infrastructure than those
targeted by Sugar.
The narrative at Nell's core attempts to provide students
pedagogical guidance missing from Sugar,
and the direct interaction model avoids keyboard
problems which have plagued Sugar's initial deployments.  Nell also
attempts to provide a greater sense of ownership through pervasive
customization.

@@

Automatic narrative generation has seen attention in
gaming, simulation, and training contexts~\cite{riedl:2008}.
We believe
Nell's synthesis of these ideas in a pedagogical context is unique.

\section{Conclusions}
We have described the design of a novel narrative direct-interface
system for education, beginning with literacy, which is personalized for and grows with
its child owner.  Personalized direct interface engages the child, and
narrative guides them through pedagogic material.  Low-cost
solar-powered hardware allows Nell to reach further into the
least-developed areas of the world to help those without traditional
educational infrastructure.
